The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Innate nature or tidal nurture?

### 2025-10-21 thesis/data.tex

Here, we use the \citepos[hereafter J+24]{jensen+2024} membership
probabilities from \emph{Gaia} data. J+24 used a Bayesian framework
incorporating proper motion (PM), colour-magnitude diagram (CMD), and
spatial information to determine the probability that a given star
belongs to the satellite or foreground/background. By accounting for PM
in particular, J+24 produced low contamination samples of candidate
member stars out to large distances from a dwarf galaxy. J+24 extended
the algorithm presented in \citet{MV2020a, MV2020b} by additionally
including a secondary, extended spatial component. J+24 detected
candidate members out to \textasciitilde10 half-light radii from the
centres of some galaxies (\(R_h\)). Similar recent work has also
included \citet{pace+li2019}; \citet{battaglia+2022};
\citet{pace+erkal+li2022}; \citet{qi+2022}.

J+24's method uses likelihoods, \({\cal L}\), representing the
probability density that a star is consistent with either the
foreground/background, \({\cal L}_{\rm bg}\), or the satellite galaxy,
\({\cal L}_{\rm sat}\). In either case, the likelihoods are the product
of a spatial, PM, and CMD component, \begin{equation}{
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
}\end{equation}

@@

For our fiducial sample, we adopt a minimum probability of
\(P_{\rm sat} = 0.2\). We do not filter on magnitudes explicitly, but
J+24's quality cuts typically only include stars with \(G < 21\). We use
the \(P_{\rm sat}\) values from the elliptical 2-component runs if a
galaxy shows evidence for an outer component, the 1-component run
otherwise. Most stars have \(P_{\rm  sat}\) values which are nearly 0 or
1, so the exact choice of probability threshold has little effect on the
resulting sample. Even at our relatively generous probability threshold
of 0.2, the purity remains high when validated against spectroscopic
line-of-sight (LOS) velocities (\textasciitilde90\%, J+24). We note,
however, that these purity estimates may be biased. Stars with LOS
velocities are typically brighter and, as a result, have more precise
\emph{Gaia} measurements. However, we find that our conclusions are
unchanged when limiting samples to only the brightest stars. Altogether,
the J+24 method provides a high-quality, low-contamination sample of
dwarf galaxy candidate member stars, which we will now investigate in
further detail.

@@

Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}
show our fiducial sample for three dSphs (Sculptor, Ursa Minor, and
Fornax), as well as the impact of different samples, in tangent-plane
coordinates, the CMD, and proper-motion space. The ``all'' sample
extends uniformly across the tangent plane, but includes a substantial
background population. The ``CMD+PM'' sample has a much lower background
density in the tangent plane, revealing the satellite more clearly. The
fiducial sample appears similar to the ``CMD+PM'' sample in the CMD and
PM planes, but excludes improbably distant stars. The ``RV members''
sample also traces out similar distributions in CMD and PM space as the
fiducial sample, with less dispersion, likely reflecting the brighter
magnitudes of stars with spectroscopic follow-up. Each selection
criterion---spatial, CMD, and PM---contributes towards a high-quality
membership assignment.

@@

To illustrate the extent of Sculptor and Ursa Minor, we highlight
distant, spectroscopically confirmed members with red-outlined stars in
Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}. In particular,
\citet{sestito+2023a}; \citet{sestito+2023b} targeted distant, bright
stars from the J+24 candidate membership list. The most distant stars
confirmed in their work lie at radii \(\gtrsim 10 R_h\) from the centres
of each dwarf. For an exponential profile, 0.05\% of stars fall outside
a radius of \(6R_h\) and less than \(1/10^6\) fall outside
\(10R_h\)---the mere presence of very distant member stars provides
evidence for deviations from an exponential profile.

@@

Fig.~\ref{fig:scl_observed_profiles} shows the derived density profiles
for Sculptor, Ursa Minor, and Fornax. We calculate density profiles for
three different samples from above: ``all'', ``CMD+PM'', and the
``fiducial'' sample. In each case, all samples coincide towards the
inner regions of the satellite. However, the density profile from
``all'' plateaus at the total background in the field at radii of 30-60
arcminutes, depending on the galaxy. By restricting stellar membership
with CMD and PM information, the background is reduced by 1--2 dex. The
``CMD+PM'' background plateau represents the density of background stars
that could be mistaken as members because of their coincident colours
and proper motions. Finally, the ``all -- background'' profile results
from subtracting the apparent background in the ``all'' profile.

While the density profiles of all samples agree in the inner regions,
they begin to deviate outside \(6R_h\). Outside this radius, the
fiducial profile extends 1--2 magnitudes below the ``CMD+PM'' selection
background. Because the fiducial profile in this region depends
critically on the spatial likelihood, its shape is vulnerable to the
assumed density profile (an exponential in J+24). We explore this in
more detail in Appendix \ref{sec:density_extra}.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/classical_dwarf_profiles.pdf}}
\caption[Classical dwarf density profiles]{The density profiles of
Sculptor, Ursa Minor, and other classical dwarfs compared to Exp2D and
Plummer density profiles. Dwarf galaxies are scaled by the half-light
radius and density at half-light radius. Sculptor and Ursa Minor have an
excess of stars in the outer regions (past \(\log R/R_h \sim 0.3\))
compared with other classical
dwarfs.}\label{fig:classical_dwarfs_densities}
\end{figure}

>>>>

Here, we use the \citepos[hereafter J+24]{jensen+2024} membership
probabilities from \emph{Gaia} data. J+24 used a Bayesian framework
incorporating proper motion (PM), colour-magnitude diagram (CMD), and
spatial information to determine the probability that a given star
belongs to the satellite or foreground/background. By accounting for PM
in particular, J+24 produced low contamination samples of candidate
member stars out to large distances from a dwarf galaxy. J+24 extended
the algorithm presented in \citet{MV2020a, MV2020b} by additionally
including a secondary, extended spatial component. J+24 detected
candidate members out to \textasciitilde10 half-light radii from the
centres of some galaxies (\(R_h\)). Similar recent work includes
\citet{pace+li2019}; \citet{battaglia+2022}; \citet{pace+erkal+li2022};
\citet{qi+2022}.

J+24's formulate membership through likelihoods, \({\cal L}\),
representing the probability density that a star is consistent with
either the foreground/background, \({\cal L}_{\rm bg}\), or the
satellite galaxy, \({\cal L}_{\rm sat}\). In either case, the
likelihoods are the product of a spatial, PM, and CMD component,
\begin{equation}{
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
}\end{equation}

@@

For our fiducial sample, we adopt a minimum probability of
\(P_{\rm sat} = 0.2\). We do not filter on magnitudes explicitly, but
J+24's quality cuts typically only include stars with \(G < 21\). We use
the \(P_{\rm sat}\) values from the elliptical 2-component runs if a
galaxy shows evidence for an outer component, the 1-component run
otherwise. Most stars have \(P_{\rm  sat}\) values which are nearly 0 or
1, so the exact choice of probability threshold has little effect on the
resulting sample. Even at our relatively generous probability threshold
of 0.2, the purity remains high when validated against spectroscopic
line-of-sight (LOS) velocities (\textasciitilde90\%, J+24).\footnote{This
  would indicate that the J+24 model probabilities are mis-calibrated.
  However, most LOS surveys of dwarf galaxies select brighter stars
  (which have better \emph{Gaia} measurements) and likely members,
  complicating the interpretation of this purity estimate.} However, we
find that our conclusions are unchanged when limiting samples to only
the brightest stars. Altogether, the J+24 method provides a
high-quality, low-contamination sample of dwarf galaxy candidate member
stars, which we will now investigate in further detail.

@@

Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}
show our fiducial sample for three dSphs (Sculptor, Ursa Minor, and
Fornax), as well as the impact of sample selection, in tangent-plane
coordinates, the CMD, and proper-motion space. The ``all'' sample
extends uniformly across the tangent plane, but includes a substantial
background population. The ``CMD+PM'' sample has a much lower background
density in the tangent plane, revealing the satellite more clearly. The
fiducial sample appears similar to the ``CMD+PM'' sample in the CMD and
PM planes, but excludes improbably distant stars. The ``RV members''
sample also traces out similar distributions in CMD and PM space as the
fiducial sample, with less dispersion, likely reflecting the brighter
magnitudes of stars with spectroscopic follow-up. Each selection
criterion---spatial, CMD, and PM---contributes towards a high-quality
membership assignment.

@@

To illustrate the extent of Sculptor and Ursa Minor, we highlight
distant, spectroscopically confirmed members with red-outlined stars in
Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}. In particular,
\citet{sestito+2023a}; \citet{sestito+2023b} targeted distant, bright
stars from the J+24 candidate membership list. The most distant stars
confirmed in their work lie at radii \(\gtrsim 10 R_h\) from the centres
of each dwarf. For an exponential profile, 0.05\% of stars fall outside
a radius of \(6R_h\) and less than \(1/10^6\) fall outside
\(10R_h\)---the mere presence of very distant member stars implies
likely deviations from an exponential profile.

@@

Fig.~\ref{fig:scl_observed_profiles} shows the derived density profiles
for Sculptor, Ursa Minor, and Fornax. We calculate density profiles for
three different samples from above: ``all,'' ``CMD+PM,'' and the
``fiducial'' sample. In each case, all samples coincide towards the
inner regions of the satellite. However, the density profile from
``all'' plateaus at the total background in the field at radii of 30-60
arcminutes, depending on the galaxy. By restricting stellar membership
with CMD and PM information, the background is reduced by 1--2 dex. The
``CMD+PM'' background plateau represents the density of background stars
that could be mistaken as members because of their coincident colours
and proper motions. Finally, the ``all -- background'' profile results
from subtracting the apparent background in the ``all'' profile.

The density profiles of all samples agree in the inner regions. However,
outside \({\sim} 6R_h\), these profiles would diverge as the
``fiducial'' sample becomes vulnerable to the assumed spatial likelihood
(an exponential in J+24). We thus truncate the fiducial profiles near
the ``CMD+PM'' background to avoid misleading density profiles. We
explore this in more detail in Appendix \ref{sec:density_extra}.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/classical_dwarf_profiles.pdf}}
\caption[Classical dwarf density profiles]{The density profiles of
Sculptor, Ursa Minor, and other classical dwarfs compared to a 2D
exponential and Plummer density profiles. Dwarf galaxies are scaled by
the half-light radius and density at half-light radius. The residuals
(lower panel) are with respect to a 2D exponential. Sculptor and Ursa
Minor have an excess of stars in the outer regions (past
\(\log R/R_h \sim 0.3\)) compared with other classical
dwarfs.}\label{fig:classical_dwarfs_densities}
\end{figure}

### 2025-10-21 thesis/discussion.tex

In the previous chapters, we have shown that Sculptor and Ursa Minor
have more extended density profiles compared to other MW classical
satellites, and that recent tides are an unlikely explanation for their
extended outer structure. We therefore suggest the reason that Scl and
UMi have extended light profiles is that they host an additional,
extended population of stars, such as a ``stellar halo.'' With our data,
other classical dwarfs lack evidence for similar extended populations.
Thus, Scl and UMi have followed a divergent evolutionary path,
motivating our discussion of extended density formation pathways here.

In this chapter, we place our results in context with past work. We
review that work, considering tidal effects on Scl and UMi and
detections of peculiarities. We then consider possible origins of
extended stellar populations in these galaxies. Scenarios include
multi-epoch star formation, past mergers depositing stars into the
outskirts, and even departures from \LCDM{}. We later discuss how future
observations and simulations may distinguish among formation scenarios.
Finally, we summarize the main conclusions of this thesis.

@@

A few studies have modelled Scl or UMi specifically, finding similar
conclusions. \citet{iorio+2019} applied idealized N-body simulations to
study tidal effects on Scl. They similarly found weak tidal effects,
even for a dark-matter-free model. Most recently,
\citet{tchiorniy+genina2025} also used idealized simulations tuned to
five classical dwarfs with a focus on the inner density, concluding that
tides do not strongly affect equilibrium assumptions or the stellar
component.

Our work extends these models, adding updated structural properties, a
broader range of orbital histories, and the influence of the LMC.
Despite these considerations, we still reach similar conclusions: tidal
effects do not shape Scl or UMi's stellar component. A non-tidal
explanation likely underlies the origin of these galaxies' outer density
excess.

@@

Besides extended stellar density profiles, Scl and UMi display other
peculiarities that may hold clues to their formation. Scl and UMi both
host at least two distinct chemodynamic populations, as revealed through
their photometric or metallicity-velocity structure\footnote{Other
  examples of galaxies with multiple populations include Carina
  \citep{battaglia+2012, fabrizio+2016, kordopatis+2016}, Fornax
  \citep{battaglia+2006, amorisco+evans2012, delpino+aparicio+hidalgo2015},
  Sextans
  \citep{battaglia+2011, cicuendez+battaglia2018, roederer+2023}, and
  Andromeda II
  \citep{mcconnachie+arimoto+irwin2007, ho+2012, delpino+2017}.}
\citep{tolstoy+2004, battaglia+2008, pace+2020}. The inner population is
younger, higher metallicity, and dynamically colder, whereas the outer
population is older, lower metallicity, and dynamically hotter.

Could these internal stellar populations map to the observed density
excess? In Scl, both populations have scale radii likely smaller than
the half-light radius. But, a third, more tentatively detected
population in \citet{arroyo-polonio+2024} would reside within the outer
density excess. In Ursa Minor, the transition between the metal-rich and
metal-poor stellar components occurs at \(\sim 30\,\)arcmin
\citep{pace+2020}, which coincides with the density excess. This
metal-poor population may thus be related to the extended stellar
density of Ursa Minor.

@@

Given their multiple stellar populations and possible inner
substructure, this hints that Scl and UMi's distinct structure today
arises from an unusual past. For example, a merger may naturally explain
both extended outer and multiple stellar populations.

@@

\textbf{Episodic star formation.} Star formation may quench and
reignite, creating successive stellar generations with differing
distributions. External star formation triggers include tidal
compression \citep{mayer+2001a, dong+lin+murray2003}, collisions with
gaseous filaments \citep{genina+2019}, perturbations from dark halos
\citep{starkenburg+helmi+sales2016}, or shocks with the MW corona
\citep{wright+2019}. More common mechanisms, like feedback or
reionization-driven quenching, may also form multiple stellar
generations
\citep{kawata+2006, benitez-llambay+2015, revaz+jablonka2018}. However,
such processes would not explain why extended stellar populations appear
to be non-universal. If an extended population was formed in an older,
induced burst of star formation, then the star formation history should
contain evidence of a corresponding burst.

\textbf{Major mergers.} Dwarf galaxy mergers may be relatively common
\citep{deason+wetzel+garrison-kimmel2014}. After a merger, stars from
the lower-mass galaxy are preferentially dispersed, forming an extended
stellar component and population gradient
\citep{benitez-llambay+2016, deason+2022}. If the galaxies contain gas,
the merger can trigger new star formation, forming a younger population
of stars \citep[e.g.,][]{genina+2019}. A few local dwarfs are suspected
to have undergone a major merger, including Tucana II, Andromeda II, and
Phoenix
\citep{lokas+2014, fouquet+2017, tarumi+yoshida+frebel2021, cardona-barrero+2021, querci+2025}.
A galaxy having undergone a major merger should harbour at least two or
three populations with distinct origin.

Just as the Milky Way's halo is believed to be built from many minor
mergers, dwarf galaxies may build accreted ``stellar halos'' composed of
even fainter satellites \citep{ricotti+polisensky+cleland2022}. Few
examples of dwarf-dwarf accretion are known, but a possible stream
around And II or chemical peculiarities around Sextans I may be such
instances \citep{amorisco+evans+vandeven2014, roederer+2023}. In this
scenario, the extended stellar populations would contain chemical
signatures from several distinct ultra-faint dwarf galaxies.

@@

\textbf{Dark-matter-free dwarfs.} Tidal dwarfs form in gas-rich tidal
streams created during the merger of two galaxies
\citep[e.g.,][]{mirabel+dottori+lutz1992, bournaud+duc2006}. Without a
dark matter halo, tidal dwarfs may be more susceptible to tides, forming
extended density profiles and inflating the velocity dispersion
\citep{casas+2012, yang+2014, wang+2024a}.

@@

We have reviewed above a number of different, possibly concurrent
explanations for the formation of extended populations. While we leave
the nature of Scl and UMi's extended stellar densities an open question,
we can discuss possible clues to different formation scenarios.

\emph{Precise chemistry}. Chemistry, particularly comparing the inner
and outer regions, could test if the halo appears to originate from a
distinct system than the dwarf. This would test if the halo is a
separate chemical population, may reveal the properties of any merged
galaxies, or show if the halo is consistent with originating from the
same galaxy.

@@

Ongoing and future observational surveys will be instrumental for
testing dwarf galaxy halo formation. Because dSph stars are usually
faint, detailed chemical analysis will rely on medium--high resolution
multi-object spectroscopy on large telescopes, such as 4MOST
\citep{skuladottir+2023} or with the extremely large telescope
\citep{jagourel+2018}. Mapping the 3D kinematics across a dwarf may
require a \emph{Gaia} successor, but ongoing surveys with JWST and HST
can derive precise proper motions within select regions of dwarf
galaxies, able to reconstruct internal structure and anisotropies
\citep[e.g.,][]{vitral+2025}. Current wide-field photometric surveys by
Rubin Observatory \citep{ivezic+2019} and the Euclid mission
\citep{euclidcollaboration+2025} will trace even fainter features around
dwarf galaxies, possibly finding halos, tidal streams, and unusual
stellar populations. Together, these observational programs promise a
new window into the detailed structure, evolution, and outskirts of
dwarf galaxies.

@@

Finally, we discuss the implications of a non-tidal density excess in
Scl and UMi. We review mechanisms ranging from episodic star formation
histories, to accretion, to departures from \LCDM{}. Future simulations
and observational programs will be able to investigate the nature of
dwarfs with ever-increasing detail. Abundant opportunities await to
uncover the cosmic origins and history of dwarf galaxies, the building
blocks of galaxies like ours.

>>>>

In the previous chapters, we have shown that Sculptor and Ursa Minor
have more extended density profiles compared to other MW classical
satellites, and that recent tides are an unlikely explanation for their
extended outer structure. We therefore suggest Scl and UMi host an
additional, extended population of stars, such as a ``stellar halo.''
With our data, other classical dwarfs lack evidence for similar extended
populations. Thus, Scl and UMi have followed a divergent evolutionary
path, motivating our following discussion of extended density formation
pathways here.

In this chapter, we compare our results to past work and
interpretations. We first review studies considering tidal effects on
Scl and UMi and detections of peculiarities. We then consider possible
origins of extended stellar populations in these galaxies. Scenarios
include multi-epoch star formation, past mergers depositing stars into
the outskirts, and even departures from \LCDM{}. We later discuss how
future studies may distinguish formation scenarios. Finally, we
summarize the main conclusions of this thesis.

@@

A few studies have modelled Sculptor or Ursa Minor specifically, finding
similar conclusions. \citet{iorio+2019} applied idealized N-body
simulations to study tidal effects on Scl. They similarly found weak
tidal effects, even for a dark-matter-free model. Most recently,
\citet{tchiorniy+genina2025} also used idealized simulations tuned to
five classical dwarfs with a focus on the inner density, concluding that
tides do not strongly affect equilibrium assumptions or the stellar
component.

Our work extends these models, adding updated structural properties, a
broader range of orbital histories, and the influence of the LMC.
Despite these considerations, we still reach similar conclusions: tidal
effects do not shape Scl or UMi's stellar component. A non-tidal process
likely underlies the origin of these galaxies' outer density excess.

@@

Besides their extended stellar density profiles, Sculptor and Ursa Minor
display other peculiarities that may hold clues to their formation. Scl
and UMi both host at least two distinct chemodynamic populations, as
revealed through their photometric or metallicity-velocity
structure\footnote{Other examples of galaxies with multiple populations
  include Carina \citep{battaglia+2012, fabrizio+2016, kordopatis+2016},
  Fornax
  \citep{battaglia+2006, amorisco+evans2012, delpino+aparicio+hidalgo2015},
  Sextans
  \citep{battaglia+2011, cicuendez+battaglia2018, roederer+2023}, and
  Andromeda II
  \citep{mcconnachie+arimoto+irwin2007, ho+2012, delpino+2017}.}
\citep{tolstoy+2004, battaglia+2008, pace+2020}. The inner population is
younger, higher metallicity, and dynamically colder, whereas the outer
population is older, lower metallicity, and dynamically hotter.

Could these internal stellar populations map to the observed density
excess? In Scl, both populations have scale radii likely smaller than
the half-light radius. However, a third, more tentatively detected
population in \citet{arroyo-polonio+2024} would reside within the outer
density excess. In Ursa Minor, the transition between the metal-rich and
metal-poor stellar components occurs at \({\sim} 30\,\)arcmin
\citep{pace+2020}, which coincides with our density excess. This
metal-poor population may thus be related to the extended stellar
density of Ursa Minor.

@@

The multiple chemical populations and possible inner structure of
Sculptor and Ursa Minor may represent the signatures of the unusual
assembly histories that produced the dwarfs' extended stellar densities.

@@

\textbf{Episodic star formation.} Star formation may quench and
reignite, creating successive stellar generations with differing
distributions. External star formation triggers include tidal
compression \citep{mayer+2001a, dong+lin+murray2003}, collisions with
gaseous filaments \citep{genina+2019}, perturbations from dark halos
\citep{starkenburg+helmi+sales2016}, or shocks with the MW corona
\citep{wright+2019}. More common mechanisms, like feedback or
reionization-driven quenching, may also form multiple stellar
generations
\citep{kawata+2006, benitez-llambay+2015, revaz+jablonka2018}. However,
such processes would not explain why extended stellar populations appear
to be non-universal. If an extended population was formed during a
separate star formation event, then the star formation history should
contain evidence of a corresponding burst.

\textbf{Major mergers.} Dwarf galaxy mergers may be relatively common
\citep{deason+wetzel+garrison-kimmel2014}. After a merger, stars from
the lower-mass galaxy are preferentially dispersed, forming an extended
stellar component and population gradient
\citep{benitez-llambay+2016, deason+2022}. If the galaxies contain gas,
the merger can also trigger new star formation, forming a younger
population of stars \citep[e.g.,][]{genina+2019}. A few local dwarfs are
suspected to have undergone a major merger, including Tucana II,
Andromeda II, and Phoenix
\citep{lokas+2014, fouquet+2017, tarumi+yoshida+frebel2021, cardona-barrero+2021, querci+2025}.
A galaxy having undergone a major merger should harbour at least two or
three populations with distinct origin.

Just as the Milky Way's halo is likely built from many minor mergers,
dwarf galaxies may have built accreted ``stellar halos'' composed of
even fainter satellites \citep{ricotti+polisensky+cleland2022}. Few
examples of dwarf-dwarf accretion are known, but a possible stream
around And II or chemical peculiarities around Sextans I may be such
instances \citep{amorisco+evans+vandeven2014, roederer+2023}. In this
scenario, the extended stellar populations would contain chemical
signatures from several distinct ultra-faint dwarf galaxies.

@@

\textbf{Dark-matter-free dwarfs.} Tidal dwarfs are galaxies that formed
in gas-rich tidal streams during the merger of two larger galaxies
\citep[e.g.,][]{mirabel+dottori+lutz1992, bournaud+duc2006}. Without a
dark matter halo, tidal dwarfs may be more susceptible to tides, forming
extended density profiles and inflating the velocity dispersion
\citep{casas+2012, yang+2014, wang+2024a}.

@@

We have reviewed a number of different, possibly concurrent explanations
for the formation of extended populations. While we leave the nature of
Scl and UMi's extended stellar densities an open question, we can
discuss possible clues to different formation scenarios.

\emph{Precise chemistry}. Chemical abundances, comparing the inner and
outer regions, can test if the halo was formed within the dwarf or
accreted. If external in origin, chemistry could reveal the properties
of any past mergers.

@@

Ongoing and future observational surveys will be instrumental for
testing dwarf galaxy halo formation. Because dSph stars are usually
faint, detailed chemical analysis will rely on medium--high resolution
multi-object spectroscopy on large telescopes, such as 4MOST
\citep{skuladottir+2023} or the extremely large telescope
\citep{jagourel+2018}. Mapping the 3D kinematics across a dwarf may
require a \emph{Gaia} successor, but ongoing surveys with JWST and HST
can derive precise proper motions within select regions of dwarf
galaxies, reconstructing internal structure and anisotropies
\citep[e.g.,][]{vitral+2025}. Current wide-field photometric surveys by
the Rubin Observatory \citep{ivezic+2019} and \emph{Euclid}
\citep{euclidcollaboration+2025} will trace even fainter features around
dwarf galaxies, possibly finding halos, tidal streams, and unusual
stellar populations. Together, these observational programs promise a
new window into the detailed structure, evolution, and outskirts of
dwarf galaxies.

@@

Finally, we discuss the implications of a non-tidal density excess in
Scl and UMi. We review mechanisms ranging from episodic star formation
histories, to accretion, to departures from \LCDM{}. Future simulations
and observational programs will be able to investigate the nature of
dwarfs with ever-increasing detail. Abundant opportunities await to
uncover the cosmic origins of dwarf galaxies, the building blocks of
galaxies like ours.

### 2025-10-21 thesis/extra_density.tex
thesis updates

To create a high-quality sample, J+24 select stars initially from Gaia
within a 2--4 degree region around each satellite with high-quality
astrometry, reliable photometry, consistent parallaxes, and broadly
consistent proper motions and colours. Stars are removed if they have
excess astrometric noise \citep[\({\rm ruwe} > 1.3\),
see][]{lindegren+2021}, colour excess \(3\sigma\) outside of
expectations \citep[from][]{riello+2021}, proper motions \(>10\,\masyr\)
in \(|\mu_{\alpha*}|\) or \(|\mu_\delta|\), magnitudes brighter than the
tip of the red giant branch\footnote{Derived from the maximum RGB
  magnitude from the associated Padova isochrone plus
  \(5\sigma\)\hspace{0pt} times the distance modulus uncertainty.} or
fainter than \(G=22\), or colours outside
\(-0.5 < G_{\rm BP} - G_{\rm RP} <  2.5\). Photometry is dereddened with
\citet{schlegel+finkbeiner+davis1998} extinction maps.

@@

\begin{figure}
\centering
\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{figures/extra_cmd_selection.png}
\caption[Colour-Magnitude sample selection]{Colour-magnitude diagram
cuts used in samples in this section. Black points represent stars
within \(1R_h\), light grey shows all field stars passing quality cuts,
and the orange polygon shows the CMD cut. The central, likely-member
stars inform the CMD selections.}\label{fig:extra_cmd}
\end{figure}

@@

@@

To bin stars, we hold fixed J+24's structural parameters and create bins
of the wider of the width \(\Delta \log R=0.05\) or containing the next
20 stars. We then derive posterior \(f_{\rm sat}\) distributions with
MCMC (48 walkers of 1000 steps each, and using the No-U-Turn sampler as
implemented in Turing.jl). The final density profile is directly derived
from \(f_{\rm sat}\) and the number of stars in each bin.

@@

Figs.~\ref{fig:mcmc_hists}, \ref{fig:mcmc_hists2} show the derived MCMC
density profiles as compared to J+24. In general, both methodologies
produce similar densities. However, J+24 tend to systematically
overestimate faint densities and confidently derive densities where the
MCMC model fails to derive a density estimate.

>>>>

To create a high-quality sample, J+24 select stars initially from Gaia
within a 2--4 degree region around each satellite with reliable
astrometry, sensible photometry, consistent parallaxes, and broadly
consistent proper motions and colours. Stars are removed if they have
excess astrometric noise \citep[\({\rm ruwe} > 1.3\),
see][]{lindegren+2021}, colour excess \(3\sigma\) outside of
expectations \citep[from][]{riello+2021}, proper motions \(>10\,\masyr\)
in \(|\mu_{\alpha*}|\) or \(|\mu_\delta|\), magnitudes brighter than the
tip of the red giant branch\footnote{Derived from the maximum RGB
  magnitude from the associated Padova isochrone plus
  \(5\sigma\)\hspace{0pt} times the distance modulus uncertainty.} or
fainter than \(G=22\), or colours outside
\(-0.5 < G_{\rm BP} - G_{\rm RP} <  2.5\). Photometry is dereddened with
\citet{schlegel+finkbeiner+davis1998} extinction maps.

@@

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/extra_cmd_selection.png}
\caption[Colour-Magnitude sample selection]{Colour-magnitude diagram
cuts used in samples in this section. Black points represent stars
within \(1R_h\), light grey shows all field stars passing quality cuts,
and the orange polygon shows the CMD cut. The central, likely-member
stars inform the CMD selections.}\label{fig:extra_cmd}
\end{figure}

@@

\textbf{Likelihood misspecification}. The J+24 likelihoods may be
misspecified. Particularly, the CMD model is simple and does not account
for, e.g., the expected number of stars on various parts of the
isochrone and assume the galaxy is a single population. In this case,
the probabilities likely do not reflect true membership probabilities.

@@

To bin stars, we hold fixed J+24's structural parameters and create bins
of width \(\Delta \log R=0.05\) or containing the next 20 stars,
whichever is wider. We then derive posterior \(f_{\rm sat}\)
distributions with MCMC (48 walkers of 1000 steps each, and using the
No-U-Turn sampler as implemented in Turing.jl). The final density
profile is directly derived from \(f_{\rm sat}\) and the number of stars
in each bin.

@@

Figs.~\ref{fig:mcmc_hists}, \ref{fig:mcmc_hists2} show the derived MCMC
density profiles as compared to J+24. In general, both methodologies
produce similar densities. However, J+24 tend to systematically
overestimate faint densities and confidently derive densities where the
MCMC model fails to determine the satellite density.
